\emph{MixFishSim}: highly resolved spatiotemporal simulations for exploring mixed fishery dynamics

### 2018-02-04 write_up/mixfishsim_ecolmod.tex
modulolinenumbers changed to discuss
Rough edit

\modulolinenumbers[5]

@@

\begin{frontmatter}

\title{A simulation framework for exploring spatio-temporal dynamics in mixed
	fisheries}

@@

\author[1]{Cóilín Minto}
\author[3]{Jean-Marc Guarini}
\author[4]{Jan-Jaap Poos}

@@

Fishing exploits spatio-temporally hetergenous fish populations without fully
selective gear which can result in unintended, unwanted catch of low quota or
protected species. Reducing these unwanted catches is crucial for biological
and economic sustainability of 'mixed fisheries' and implementation of an
ecosystem approach to fishing. \\

In order to implement effective spatial measures to reduce discards there needs
to be a good understanding of spatio-temporal fishery dynamics. However,
scientific advice is limited by a lack of knowledge of population distribution,
movement and how fishers interact with different fish populations. This is
because data on location of fish at high temporal and spatial resolutions is
expensive and difficult to collect and therefore proxy inferred from either
scientific survey or commercial catch data is often used to model
distributions, but often limited in its spatial or temporal resolution. \\

We develop a highly resolved spatio-temporal simulation model that incorporate
i) delay-difference population dynamics, ii) population movement dynamics using
Gaussian Random Fields to simulate patchy, hetergenously distributed
populations, iii) fishery dynamics for multiple fleet characteristics based on
a correlated random walk and learned behaviour. \\

Using our model we simulation 20 years of exploitation of the fish populations
and use the results from the fishing model to draw inference on the underlying
population structures. We compare this inference to i) a simulated fixed-site
sampling design commonly used for fisheries monitoring purposes, ii) the
underlying population structures input to the simulation, to establish whether
fishing dependent data, which is obtained from a biased sampling due to fishers
targeting, provides a robust picture of spatio-temporal distributions. [We
simulate a closure based on areas defined from commercial catch data and assess
its effectiveness on reducing catches of a fish population.] \\

@@

Fishers exploit fish populations that are heterogenously distributed in space
and time without prior knowledge of species distributions using non-selective
fishing gear. Fisheries that catch an assemblage of species, known as mixed
fisheries, when managed by single-species quotas can end up discarding
overquota catch leading to overexploitation of fish populations. Reducing
discarding is crucial to ensure biological and economic sustainability of
fisheries and implementation of an ecosystem approach to fisheries. As such
there is increasing interest in technical solutions such as gear, spatial
closures as a way of avoiding discards.  \\

Use of spatial management as a tool has been proposed as a method to reduce
discards. However, its implementation is hampered by lack of knowledge of fish
and fishery spatiotemporal dyanmics and understanding of the scale at which
processes are important for management. Understanding of the correct scale for
spatial management is crucial in order to implement measures at a resolution
that ensures effective management\cite{Dunn2016} while minimising economic
impact; for example a scale which promote species avoidance for vulnerable or
low quota species while allowing continance of sustainable fisheries for
available quota species. \\

Ensuring measures are implemented at an appropriate scale has been a challenge
in the a past that has led to ineffectual measures with unitended consequences
such as limited impact towards the management objective or increased benthic
impact on previously unexploited areas (e.g. the cod closure in the North
Sea\cite{Rijnsdorp2001,Dinmore2003}). Since then more refined spatial
information has become available through the combination of logbook and Vessel
Monitoring System (VMS) data\cite{Lee2010, Bastardie2010, Gerritsen2012,
	Mateo2016} and more real-time spatial management has been able to be
implemented (e.g. \cite{Holmes2011}), though such information is patchy and
derived from an inherently biased sampling programme (i.e. targeted fishing).
Further, fishers generally only recorded landings (not catch) on a daily basis.
This leads to questions about the validity of inferences that can be drawn from
landings data assigned to VMS activity pings. \\

In order to test the assumption that VMS associated landings data can be used
to draw inference on the underlying population structures we develop a
simulation model where population dynamics are known rather than inferred from
sampling or commercial catches. Population movement is driven by a random
(diffusive) and directed (advective) process and we incorporate
characterisation of a number of different fisheries exploiting four fish
populations with different spatial and population demographics.\\

Using our model we simulation 20 years of exploitation of the fish populations
and use the results from the fishing model to draw inference on the underlying
population structures.  We compare this inference to i) a simulated
[stratified] fixed-site sampling design commonly used for fisheries monitoring
purposes, ii) the underlying population structures input to the simulation.\\

@@

We develop a simulation model with a modular event-based approach, where
modules are implemented on independent time-scales appropriate to capture the
characteristic of the process modelled. The fishing model operated on a
tow-by-tow basis, while population dynamics (fishing and natural mortality,
growth) operate on a daily time-step.  Population movement occurs on a
weekly time-step, while recruitment occurs periodcally each year for a set
time period (e.g. 3 weeks) at at specified point individual to a species. The
simulation framework is implemented in the statistical software package R
\cite{RCoreTeam2017}; available as an R package from the authors github
(\url{www.github.com/pdolder/MixFishSim}). The overall model structure is
summerised in Figure 1.\\

In the following section, we describe each of the model components; 1)
Population dynamics, 2) Recruitment dynamics, 3) Population movement, 4)
fishery dynamics.

@@

% -------------------------------------------------
% A last flourish which breaks all the rules
\draw [->,purple, dotted, thick, shorten >=1mm]
  (p3.south) -- ++(5mm,-3mm)  -- ++(87mm,0mm) 
  |- node [black, near end, yshift=1.75em, it]
    {} ($(p1.north) + (0,0.3)$);
% -------------------------------------------------
\end{tikzpicture}
\caption{Overview Schematic of simulation model}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

where $C$ is the summed catch from the fishing model across all fleets and
vessels for the population during the day, and $B$ the daily biomass for the
species. \\

@@

Recruitment is modelled through a function relating the biomass at time of
recruitment to recruits. In the simulation, it can be modelled either either as
a stochastic Beverton-Holt stock-recruit form (\cite{Beverton1957}):

\begin{equation*}
	\begin{split}
	\bar{R} = & \frac{(\alpha * B)}{(\beta + B)} \\
	     R \sim & N[(\bar{R},\sigma^2)]
	\end{split}
\end{equation*}

Where $\alpha$ is the maximum recruitment rate, $\beta$ the spawning stock
biomass (SSB) required to produce half the maximum, and $B$ current SSB; \\

@@

In order to simulate how fish populations might be distributed in space and
time, we employed a Gaussian spatial process to model habitat suitability for
each of the populations, with an advection-diffusion process to control how the
populations moved over time. \\

@@

Where $d_{AB}$ is the euclidean distance between cell $A$ and cell $B$, and
$\lambda$ is a given rate of decay.\\

@@

Several studies (e.g.\cite{Hutton2004, Tidd2012, Girardin2015}) have confirmed
past activity and past catch rates are strong predictors of fishing location
choice. For this reason, the fleet dynamics sub-model includes a learning
component, where a vessels initial fishing location in a trip is based on
selecting from previously successful fishing locations. This is achieved by
sorting all previous fishing events in the previous trip as well as the
previous time periods in past years, and choosing randomly from the top x \% of
fishing events in value.  Simulation testing indicating this increased the mean
value of catches for the vessels, over just relying on the correlated random
walk function.

@@

Fishing locations within a trip are determined by a Random Walk process. A
Random walk was chosen as it is commonly used in ecology to describe animal
movement which searching for homogeneously distributed prey about which there
is uncertain knowledge. In a random walk, movement is a stochastic process
through a series of steps that can either be equal in length or take some other
functional form.  The direction of the random walk can be correlated, a
characteristic known as 'persistence', providing some overall location
of directional movement \cite{Codling2008} or uncorrelated. \\

@@

We use a modified random walk where directional change is based on a correlated
circular distribution where a favourable fishing ground is likely to be "fished
back over" by the vessel returning in the direction it came from and step
length (i.e. the distance travelled from the current to the next fishing
location) is determined by relating recent fishing success, measured as the
summed value of fish caught, $$Rev = \sum_{s=1}^{\infty} C_{s} \cdot Pr_{s}$$
where $C_{s}$ is catch of a species, and $Pr_{s}$ price of a species, to step
distance. Here, when fishing is successful vessels remain in a similar location
and continue to exploit the local fishing grounds. When unsuccessful, they move
some distance away from the current fishing location. The movement distance
retains some degree of stochasticity, which can be controlled separately.

@@

with $k$ the concentration parameter from the von mises distribution which we
correlate with the revenue so that $k = (Rev + 1 / RefRev) * max_{k}$,
where $max_{k}$ is the maximum concentration value, $k$, and RefRev is
parameterised as for $\beta_{3}$ in the step length function.

@@

A fisheries-independent survey is simulated where fishing on a regular grid
begins each year at the same time for a given number of stations (a fixed
station survey design). Catches of the populations present are recorded but not
removed from the population. This provides a fishery independent snap shot of
the populations at a regular spatial distribution each year, similar to
scientific surveys undertaken by fisheries research agencies.

@@

We set up with simulation to run for 20 years based on a 100 X 100 square grid,
with five fleets of 20 vessels each and four fish populations. Fishing takes
place four times a day per vessel and five days a week, while population
movement is every week.

@@

The fleets were parameterised to reflect five different characteristics based
on targeting preference and exploitation dynamics (Table \ref{tab:2}). This
ensures that different fleets have different spatial dynamics, preferentially
targeted different fish populations. The stochasticity in the random walk
process ensures that different vessels within a fleet have slightly different
spatial distributions based on individual experience, while the step function
was parameterised dynamically so that vessels take smaller steps where the
fishing location yields in the top X Quartile of the value available in that
year (as defined per fleet in Table \ref{tab:2}). \\

@@

\begin{itemize}
	\item Spatial dynamics: e.g. Figure 14. showing the population movement
		across weeks, including a spawning period.
	\item Overall population trends: e.g. Figure 3, showing the population
		dynamics at play.
	\item Realised step function: e.g. Figure 13, showing how the movement
		and turning angle responds. This could be combined with Figure
		7 for a more complete visualisaiton.
	\item Catch composition and spatial clustering: e.g. Figure 10, showing
		the importance of spatial resolution.
	\item Some measure of temporal trends/changes in spatial composition -
		could pick a certain cell and show how the proportions change
		over time ? An example is Figure 15, though there is little
		variation here. I think this is because the static habitat with
		population movement is far more stable than the previous
		directly affected population distributions (where there was an
		advective-diffusive process from the SPDE). Might need to
		consider how to change this?  Could include a spatio-temporally
		changing habitat suitability covariate, e.g. temperature ??
	\item Comparison of population structure from i) commercial sampling,
		ii) fisheries independent survey, iii) real population. This
		should be some statistical measure...not sure best approach
		here.
	\item [If we perform the closure, some measure of cost/benefit based on
		different resolutions of data and data used (i.e. commercial,
		survey, real populations). This could be on the basis of
		closing x cells, giving cell references to the model so they
		cannot be fished. Need to consider whether this is also a
		seasonal closure.]
\end{itemize}

>>>>

\modulolinenumbers[1]

@@

\begin{frontmatter}
\title{\emph{MixFishSim}: highly resolved spatiotemporal simulations for exploring mixed fishery dynamics}

%%\title{A simulation framework for exploring spatio-temporal dynamics in mixed
%%	fisheries}

@@

\author[1]{Cóilín Minto}
\author[3]{Jean-Marc Guarini}
\author[4]{Jan Jaap Poos}

@@

Fishing exploits spatially and temporally hetergenous fish populations, using species-unselective 
gear that can result in unintended, unwanted catch of low quota or
protected species. Reducing these unwanted catches is crucial for biological
and economic sustainability of `mixed fisheries' and implementation of an
ecosystem approach to fishing. \\

To implement effective spatial measures to reduce discards 
a good understanding of spatio-temporal fishery dynamics is required. However,
traditional scientific advice is limited by a lack of highly resolved knowledge of population distribution,
movement and how fishers interact with different fish populations. This reflects that 
data on fish location at high temporal and spatial resolutions is
expensive and difficult to collect and therefore proxies inferred from either
scientific surveys or commercial catch data are often used to model
distributions, often with limited spatial and temporal resolution. \\

To understand how resolution impacts mixed fisheries inference, we develop a highly 
resolved spatio-temporal simulation model incorporating:
i) delay-difference population dynamics, ii) population movement using
Gaussian Random Fields to simulate patchy, hetergenously distributed
populations, and iii) fishery dynamics for multiple fleet characteristics based on
targetting via correlated random walk movement and learned behaviour. \\

We simulate 20 years of exploitation of the fish populations
and use the results from the fishing model to draw inference on the underlying
population structures. We compare this inference to i) a simulated fixed-site
sampling design commonly used for fisheries monitoring purposes, and ii) the
true underlying population structures input to the simulation, to establish the potential and limitations of
fishery-dependent data - an inherently biased sampling method due to fisher's
targeting- to provide a robust picture of spatio-temporal distributions. [We
simulate a closure based on areas defined from commercial catch data and assess
its effectiveness on reducing catches of a fish population.] \\

@@

Fishers exploit fish populations that are heterogenously distributed in space
and time with verying knowledge of species distributions using species-unselective
fishing gear. Fisheries that catch an assemblage of species, known as mixed
fisheries, when managed by single-species quotas can end up discarding
overquota catch leading to overexploitation of fish populations. Reducing
discarding is crucial to ensure biological and economic sustainability of
fisheries and implementation of an ecosystem approach to fisheries. As such
there is increasing interest in technical solutions such as gear and spatial
closures as ways of avoiding discards.  \\

Use of spatial management as a tool has been proposed as a method to reduce
discards. However, its implementation is hampered by lack of knowledge of fish
and fishery spatiotemporal \textcolor{red}{[use this spelling throughout]} dynamics and understanding of the scale at which
processes are important for management. Understanding the correct scale for
spatial management is crucial in order to implement measures at a resolution
that ensures effective management\cite{Dunn2016} while minimising economic
impact. For example, a scale that promotes species avoidance for vulnerable or
low quota species while allowing continuance of sustainable fisheries for
available quota species. \\

Ensuring measures are implemented at an appropriate scale has been a challenge
in the past that has led to ineffectual measures with unitended consequences
such as limited impact towards the management objective or increased benthic
impact on previously unexploited areas (e.g. the cod closure in the North
Sea\cite{Rijnsdorp2001,Dinmore2003}). Since then more refined spatial
information has become available through the combination of logbook and Vessel
Monitoring System (VMS) data\cite{Lee2010, Bastardie2010, Gerritsen2012,
	Mateo2016} and more real-time spatial management has been possible 
	(e.g. \cite{Holmes2011}). Such information is, however, patchy and
derived from an inherently biased sampling programme (i.e. targeted fishing).
Further, fishers generally only recorded landings (not catch) on a daily basis.
This leads to questions about the validity of inference that can be drawn from
landings data assigned to VMS activity pings. \\

In order to understand challenges that face VMS-linked landings 
to draw inference on the underlying population structure we develop a
simulation model where population dynamics are highly-resolved in space and time and are 
known rather than inferred from
sampling or commercial catches. Population movement is driven by a random
(diffusive) and directed (advective) process and we incorporate
characterisation of a number of different fisheries exploiting four fish
populations with different spatial and population demographics.\\

Using our model we simulate 20 years of exploitation of the fish populations
and use the results from the fishing model to draw inference on the underlying
population structures.  We compare this inference to: i) a stratified fixed-site sampling survey 
design commonly used for fisheries monitoring
purposes \textcolor{red}{[not sure if this is the fisheries-independent survey?]} and ii) the underlying population structures input to the simulation.\\

@@

We develop a simulation model with a modular event-based approach, where
modules are implemented on independent time-scales appropriate to capture the
characteristic of the process modelled (Figure 1). The fishing model operated on a
tow-by-tow basis, while population dynamics (fishing and natural mortality,
growth) operate on a daily time-step.  Population movement occurs on a
weekly time-step, while recruitment occurs periodcally each year for a set
time period (e.g. 3 weeks) at at specified point individual to a species. The
simulation framework is implemented in the statistical software package R
\cite{RCoreTeam2017}; available as an R package from the authors github
(\url{www.github.com/pdolder/MixFishSim}).\\

Here we describe each of the model components; 1)
Population dynamics, 2) Recruitment dynamics, 3) Population movement, 4)
fishery dynamics.

@@

% -------------------------------------------------
% A last flourish which breaks all the rules
\draw [->,purple, dotted, thick, shorten >=1mm]
  (p3.south) -- ++(5mm,-3mm)  -- ++(87mm,0mm) 
  |- node [black, near end, yshift=1.75em, it]
    {} ($(p1.north) + (0,0.3)$);
% -------------------------------------------------
\end{tikzpicture}
\caption{Overview Schematic of simulation model \textcolor{red}{Make a table with all variables and symbols so that you can reference them in the caption. Expand the caption to give as much standalone info as possible, e.g., what's blue, what's green, arrow colours, etc.}}
\label{fig:overview}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

where $C$ is the summed catch from the fishing model across all fleets and
vessels for the population during the day, and $B$ the daily biomass for the
species. \textcolor{red}{[link $F$ to effort and catchability - as I think we have F as an emergent property of the fleets rather than something we solve for (I could be wrong though!)]}\\

@@

Recruitment is modelled through a function relating the mature biomass to recruits at time of
recruitment. In \emph{mixfishsim}, it can be modelled either either as
a stochastic Beverton-Holt stock-recruit form (\cite{Beverton1957}):

\begin{equation*}
	\begin{split}
	\bar{R} = & \frac{(\alpha * B)}{(\beta + B)} \\
	     R \sim & N[(\bar{R},\sigma^2)]
	\end{split}
\end{equation*}
\textcolor{red}{[better to use lognormal variability to avoid negative recruitment events at low biomasses.]}
Where $\alpha$ is the maximum recruitment rate, $\beta$ the spawning stock
biomass (SSB) required to produce half the maximum, and $B$ current SSB; \\

@@

To simulate how fish populations might be distributed in space and
time, we employed a Gaussian spatial process to model habitat suitability for
each of the populations, with an advection-diffusion process to control how the
populations moved over time. \textcolor{red}{[say why - balance between realism and practicalities of IBMs for the fish population]}\\

@@

Where $d_{AB}$ is the euclidean distance between cell $A$ and cell $B$, and
$\lambda$ is a given rate of decay. \textcolor{red}{[this is great but do explain $H_{ab}$ and all variables immediately on presentation throughout.]}\\

@@

Several studies (e.g.\cite{Hutton2004, Tidd2012, Girardin2015}) have confirmed
past activity and past catch rates are strong predictors of fishing location
choice. For this reason, the fleet dynamics sub-model includes a learning
component, where a vessel's initial fishing location in a trip is based on
selecting from previously successful fishing locations. This is achieved by
sorting all previous fishing events in the previous trip as well as the
previous time periods in past years, and choosing randomly from the top x \% of
fishing events in value.  Simulation testing indicated that this learning increased the mean
value of catches for the vessels, over just relying on the correlated random
walk function.

@@

Fishing locations within a trip are determined by a modified random walk process. A
random walk type was chosen as it is the simplest assumption commonly used in ecology to describe animal
movement which searching for homogeneously distributed prey about which there
is uncertain knowledge. In a random walk, movement is a stochastic process
through a series of steps that can either be equal in length or take some other
functional form.  The direction of the random walk can be correlated, a
characteristic known as `persistence', providing some overall location
of directional movement \cite{Codling2008} or uncorrelated. \\

@@

We use a modified random walk where directional change is based on a correlated
circular distribution where a favourable fishing ground is likely to be ``fished
back over" by the vessel returning in the direction it came from and step
length (i.e. the distance travelled from the current to the next fishing
location) is determined by relating recent fishing success, measured as the
summed value of fish caught, $$Rev = \sum_{s=1}^{\infty} C_{s} \cdot Pr_{s}$$
where $C_{s}$ is catch of a species, and $Pr_{s}$ price of a species, to step
distance. Here, when fishing is successful vessels remain in a similar location
and continue to exploit the local fishing grounds. When unsuccessful, they move
some distance away from the current fishing location. The movement distance
retains some degree of stochasticity, which can be controlled separately.

@@

with $k$ the concentration parameter from the von mises distribution which we
correlate with the revenue so that $k = (Rev + 1 / RefRev) * max_{k}$,
where $max_{k}$ is the maximum concentration value, $k$, and RefRev is
parameterised as for $\beta_{3}$ in the step length function \textcolor{red}{[wonderful!]}.

@@

A fisheries-independent survey is simulated where fishing on a regular grid
begins each year at the same time for a given number of stations (a fixed
station survey design). Catches of the populations present are recorded but not
removed from the population. This provides a fishery independent snapshot of
the populations at a regular spatial distribution each year, similar to
scientific surveys undertaken by fisheries research agencies.

@@

To illustrate the capabilities on \emph{MixFishSim}, we investigate the influence of ... \textcolor{red}{[expand]}. To do so, we first set up with simulation to run for 20 years based on a 100 X 100 square grid,
with five fleets of 20 vessels each and four fish populations. Fishing takes
place four times a day per vessel and five days a week, while population
movement is every week.

@@

The fleets were parameterised to reflect five different characteristics based
on targeting preference and exploitation dynamics (Table \ref{tab:2}). This
ensures that different fleets have different spatial dynamics, preferentially
targeted different fish populations. The stochasticity in the random walk
process ensures that different vessels within a fleet have slightly different
spatial distributions based on individual experience, while the step function
was parameterised dynamically so that vessels take smaller steps where the
fishing location yields in the top X \textcolor{red}{[??]}quartile of the value available in that
year (as defined per fleet in Table \ref{tab:2}). \\

@@

\begin{itemize}
	\item Spatial dynamics: e.g. Figure 14. showing the population movement
		across weeks, including a spawning period.
	\item Overall population trends: e.g. Figure 3, showing the population
		dynamics at play.
	\item Realised step function: e.g. Figure 13, showing how the movement
		and turning angle responds. This could be combined with Figure
		7 for a more complete visualisaiton.
	\item Catch composition and spatial clustering: e.g. Figure 10, showing
		the importance of spatial resolution.
	\item Some measure of temporal trends/changes in spatial composition -
		could pick a certain cell and show how the proportions change
		over time ? An example is Figure 15, though there is little
		variation here. I think this is because the static habitat with
		population movement is far more stable than the previous
		directly affected population distributions (where there was an
		advective-diffusive process from the SPDE). Might need to
		consider how to change this?  Could include a spatio-temporally
		changing habitat suitability covariate, e.g. temperature ?? \textcolor{red}{[temperature would be very interesting and cover varying spatial fields]}
	\item Comparison of population structure from i) commercial sampling,
		ii) fisheries independent survey, iii) real population. This
		should be some statistical measure...not sure best approach
		here.
	\item [If we perform the closure, some measure of cost/benefit based on
		different resolutions of data and data used (i.e. commercial,
		survey, real populations). This could be on the basis of
		closing x cells, giving cell references to the model so they
		cannot be fished. Need to consider whether this is also a
		seasonal closure.]
\end{itemize}
